High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-04-14 PNAS_manuscript/PNAS_manuscript.tex

In this study, we target the high-purity anorthosite xenoliths of the Beaver River diabase in the Midcontinent Rift. While magmatic activity within the Midcontinent Rift was protracted, there were intervals of particularly rapid volcanism and voluminous emplacement of intrusions \cite{Swanson-Hysell2021a}. The ca. 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the main stage of Midcontinent Rift activity. The magma that formed the 1091.7$\pm$0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. These anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust---an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}. They are attractive targets for paleomagnetic study as plagioclase crystals can protect magnetic inclusions from alteration. In addition, the alteration of the plagioclase crystals does not readily result in the formation of secondary iron oxides in contrast with Fe-silicate minerals such as olivine and pyroxene. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal (shallowly emplaced) intrusions of the Beaver River diabase \cite{Zhang2021b}. They would have been heated to tholeiitic magma temperature which at $\sim$1100\textdegree C is below the melting point of the xenolith plagioclase (70$\%$ anorthite), but well above the Curie temperature of magnetite \cite{Zhang2021b}. Fe-Ti oxides within the plagioclase likely exsolved above magnetite Curie temperature \cite{Bian2021a} and subsequently cooled and acquired thermal remanent magnetizations in conjunction with the host diabase at a paleolatitude of 22 $\pm$ 2\textdegree$\;$(calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a}). Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. The anorthosite xenoliths show low anisotropy of thermal remanent magnetization acquisition and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows that the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals without strong preferred orientations. Thermal modeling results and paleomagnetic directional data show that the anorthosite xenoliths acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetization data show the anorthosite xenoliths to have dominantly single-component magnetizations that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C, consistent with remanence being held by low-titanium titanomagnetite \cite{Zhang2021b}.

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM_full.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F,G,H) of an anorthosite sample from the Beaver River anorthosite xenolith in the Silver Bay region from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and from a distinct anorthosite within the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture of crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Maps of the vertical component of magnetic field (B$_z$) developed with a quantum diamond microscope (QDM) show relatively weak magnetic sources within plagioclase crystals in the Beaver River anorthosite (in E) relative to the strongly magnetic large oxide needles within Duluth Complex plagioclase (in F). The B$_z$ color scale is an order of magnitude greater in the Duluth Complex anorthosite maps (F, H) than the maps for the Beaver River anorthosite (E, G). Panels (E) to (G) and (F) to (H) show experiments performed on both samples where we apply a first field along the y axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The remanent magnetizations acquired by both anorthosites were mapped after the application of each field. The magnetic images show that remanent magnetizations of Beaver River anorthosite (i.e. the individual dipoles visible with paired red +B$_z$ and blue -B$_z$ lobes) align well with the first applied field direction (E) and then rotated to align with the second applied field direction indicating minimal anisotropic behavior (G). In contrast, the remanence magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 acquired an oblique remanence direction with respect to the field direction (F). After the application of an orthogonal 300 mT external field, magnetization of those needles in plagioclase 1 did not change direction due to strong shape anisotropy whereas the magnetization of needles in plagioclase 2 flipped, but with the acquired remanence remaining oblique to the field direction (H). Insets in (E) to (H) show example dipoles direction responses to the application of orthogonal IRMs. This experiment illustrates that the Beaver River anorthosite xenolith has much lower magnetic anisotropy than the Duluth anorthosite xenolith. The unit of the axes in the QDM maps are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

>>>>

In this study, we target the high-purity anorthosite xenoliths of the Beaver River diabase in the Midcontinent Rift. While magmatic activity within the Midcontinent Rift was protracted, there were intervals of particularly rapid volcanism and voluminous emplacement of intrusions \cite{Swanson-Hysell2021a}. The ca. 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the main stage of Midcontinent Rift activity. The magma that formed the 1091.7$\pm$0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. These anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust---an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}. They are attractive targets for paleomagnetic study as plagioclase crystals can protect magnetic inclusions from alteration. In addition, the alteration of the plagioclase crystals does not readily result in the formation of secondary iron oxides in contrast with Fe-silicate minerals such as olivine and pyroxene. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal (shallowly emplaced) intrusions of the Beaver River diabase \cite{Zhang2021b}. They would have been heated to tholeiitic magma temperature ($\sim$1100\textdegree C) which is below the melting point of the xenolith plagioclase (70$\%$ anorthite), but well above the Curie temperature of magnetite \cite{Zhang2021b}. Fe-Ti oxides within the plagioclase likely exsolved above magnetite Curie temperature \cite{Bian2021a} and subsequently cooled and acquired thermal remanent magnetizations in conjunction with the host diabase at a paleolatitude of 22 $\pm$ 2\textdegree$\;$(calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a}). Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. The anorthosite xenoliths show low anisotropy of thermal remanent magnetization acquisition and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows that the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals without strong preferred orientations. Thermal modeling results and paleomagnetic directional data show that the anorthosite xenoliths acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetization data show the anorthosite xenoliths to have dominantly single-component magnetizations that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C, consistent with remanence being held by low-titanium titanomagnetite \cite{Zhang2021b}.

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=12.4 cm]{Petro_QDM_full.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F,G,H) of an anorthosite sample from the Beaver River anorthosite xenolith in the Silver Bay region from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and from a distinct anorthosite within the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture of crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Maps of the vertical component of magnetic field (B$_z$) developed with a quantum diamond microscope (QDM) show relatively weak magnetic sources within plagioclase crystals in the Beaver River anorthosite (in E) relative to the strongly magnetic large oxide needles within Duluth Complex plagioclase (in F). The B$_z$ color scale is an order of magnitude greater in the Duluth Complex anorthosite maps (F, H) than the maps for the Beaver River anorthosite (E, G). Panels (E) to (G) and (F) to (H) show experiments performed on both samples where we apply a first field along the y axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The remanent magnetizations acquired by both anorthosites were mapped after the application of each field. The magnetic images show that remanent magnetizations of Beaver River anorthosite (i.e. the individual dipoles visible with paired red +B$_z$ and blue -B$_z$ lobes) align well with the first applied field direction (E) and then rotated to align with the second applied field direction indicating minimal anisotropic behavior (G). In contrast, the remanent magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 acquired an oblique remanence direction with respect to the field direction (F). After the application of an orthogonal 300 mT external field, magnetization of those needles in plagioclase 1 did not change direction due to strong shape anisotropy whereas the magnetization of needles in plagioclase 2 flipped, but with the acquired remanence remaining oblique to the field direction (H). Insets in (E) to (H) show example dipole directions (fit using the algorithm of \cite{Lima2016a}) in response to the application of orthogonal IRMs. The directional changes of the Beaver River anorthosite xenolith sources indicate a lower magnetic anisotropy compared to the relative lack of change for the Duluth Complex anorthosite xenolith sources. The unit of the axes in the QDM maps are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}
